---
title: "Non-response in useR! 2016 Survey"
author: Heather Turner and Julie Josse
date: "7 October 2016"
output: html_document
type: report
---

```{r setup, include=FALSE}
library(ggplot2)
library(knitr)
library(scales)
library(dplyr)
library(tidyr)
library(forcats)
library(matrixStats)
load("../../../survey/survey.RData")
## From Julie's genderizer + manual gender assignment
gender <- data_frame(Gender = c("Female", "Male", "Undetermined"), 
                     Freq = c(235, 593, 76))
opts_chunk$set(echo = FALSE, message = FALSE, fig.align = "center")
```

## Background

<!-- Survey report says 456 responses, what are extra 2? -->
The useR! survey received a total of `r sum(!rowAlls(is.na(survey_data)))` responses out of 
`r sum(gender$Freq)` attendees. Anticipating a high proportion of non-response,
questions were included in the survey on participants' gender, registration type
and country of residence, with a view to adjusting observed counts in each 
cross-classified group to match the proportions in the total useR! 2016 
population. Unfortunately the country of residence was not recorded during
registration and could not be reliably derived from available information such
as affiliation. Although registration type should have been available,
in practice it was missing for about one third of participants. This leaves
gender, which was (self-)recorded for presenters, but not for other attendees. 
In order to obtain a gender assignment for all participants, the first names
were analysed with the **genderizer** package, which predicts gender from first 
name, based on profiles scraped from the web. For participants where the gender
was unknown, the genderizer prediction was used if it was based on a sample > 10
and had probability > 0.9, otherwise the gender was assigned based on a manual
web search, where this led to a clear assignment. This gave estimates of 
`r with(gender, Freq[Gender == "Female"])` female, 
`r with(gender, Freq[Gender == "Male"])` male and 
`r with(gender, Freq[Gender == "Undetermined"])` undetermined.

Assuming that the gender distribution is similar in the undetermined group, the 
survey population overrepresents female attendees by about 10%:

```{r gender_breakdown}
obsGender <- as_data_frame(table(survey_data$Q2))
allGender <- full_join(obsGender, filter(gender, Gender != "Undetermined"), 
                       by = c("Var1" = "Gender")) %>%
  mutate(Freq = 100 * Freq/sum(Freq, na.rm = TRUE),
         n = 100 * n/sum(n, na.rm = TRUE))
names(allGender) <- c("Gender", "Survey Percentage", "Attendee Percentage")
kable(allGender, digits = 2)
```

(The assumption that the undetermined have a similar gender distribution seems
reasonable, as similar estimates of attendee breakdown were obtained based on
genderizer results only, where there were 
`r round(100 * with(attendee_gender, sum(n[is.na(gender)])/sum(n)), 2)`%
undetermined). This over-representation of females is a common phenomenon in 
surveys and particularly unsurprising here as females may be more motivated to 
respond to a survey run by the Women in R taskforce.

```{r est_weights}
# only considered Male and Female categories in sample and pop
allMF <- inner_join(gender, obsGender, by = c("Gender" = "Var1"))
weight <- mutate(allMF,
       w = (Freq/sum(Freq)/(n/sum(n))))
# add weights to survey data - set weight for non-binary to be 1
survey_data <- left_join(survey_data, select(weight, Gender, w), by = c("Q2" = "Gender")) %>%
  replace_na(list(w = 1))
```

## Adjusting for non-response

In order to adjust for the difference in gender balance between the sample and 
the full population, weights may be computed for males and females so that the 
ratio of females to males based on adjusted counts (sum of weights in each 
category) is equal to that estimated for the full sample. Specifically, the 
weight for females is 
`r with(weight, round(w[Gender == "Female"], 2))` 
and the weight for males is 
`r with(weight, round(w[Gender == "Male"], 2))`.

These weights can then be used in further analysis, for example when tabulating 
by another survey variable. Here we will consider a few examples for 
illustration.

### Age

The unweighted and weighted densities have a very similar shape, but using the
weighting shifts the density slightly to the left, i.e. when females are 
down-weighted there is a slight shift to a generally older population.

```{r}
# get data on age, omit missing
data <- na.omit(select(survey_data, Q3, w))
# show unweighted (w = 1) and weighted
data <- bind_rows(data_frame(Q3 = data$Q3,
                             w = 1),
                  data,
                  .id = "Weight")
# need to make weights sum to 1 to for true density plot
data <- data %>%
  group_by(Weight) %>%
  mutate(w = w/sum(w))

ggplot(data, aes(x = Q3, weight = w, fill = Weight, group = Weight)) + 
         geom_density(alpha = 0.5, color = NA) +
  labs(x = "Year of Birth", y = "Density",
       title = paste0("Year of birth of useR! attendees"),
       fill = NULL) +
  theme_minimal(base_family = "RobotoCondensed-Regular") +
  scale_fill_discrete(labels = c("unweighted", "weighted"))
  
```

### Education level

Here there is very little difference between the unweighted and weighted counts.

```{r Q7}
xtab <- function(...){
  tab <- survey_data %>%
    group_by_(...) %>%
    summarise(Count = n(),
              Adj.Count = sum(w))
  kable(tab, digits = 2)
}

survey_data <- survey_data %>%
  mutate(Q7 = factor(Q7, levels = c("Professional degree or certification",
                                      "Doctorate", "Masters degree",
                                      "Undergraduate degree",
                                      "High school/Secondary school or lower")))

xtab("Q7")
```

### Employment status

Again the differences between unweighted and weighted counts are small here,
however there is a consistent shift from less financially secure positions 
(student, unemployed and employed non-permenantly in academia) to more 
financially secure positions (employed permenantly in academia, employed in 
industry). This tallies with the younger distribution of females noted
above.

```{r Q8}
survey_data <- survey_data %>%
  mutate(Q8 = factor(Q8, levels = c("Employed in industry", 
                                    "Employed permanently in academia", 
                                    "Employed non-permanently in academia",
                                    "Employed in government/non-profit", 
                                    "Self-employed", "Retired",
                                    "Student", "Unemployed")))
xtab("Q8")
```

### Time using R

With the weighted counts we see a slight shift towards greater experience; this
again tallies with females being generally younger and less advanced in their 
careers.

```{r Q11}
survey_data <- survey_data %>%
  mutate(Q11 = factor(Q11, levels = c("> 10 years", "5-10 years", "2-5 years",
                                      "1-2 years", "6-12 months", "< 6 months")))
xtab("Q11")
```

### Recreational use of R

With the weighted counts we see a shift towards using R for recreational 
activity alone or as well as for work. Since the weighting down-weights 
females relative to males this suggests males are more able or more willing to 
use R for fun.

```{r Q14}
xtab("Q14")
```

### Summary

The examples above illustrate that the difference between unweighted and 
weighted counts tends to be small and is unlikely to affect conclusions
drawn about the distribution of a particular variable in the useR! population as
a whole.

Where there is a material difference between unweighted and weighted counts,
this can be attributed to gender, since the weighting only depends on this 
factor. In this case it would be more interesting to explore the gender 
difference by showing a summary grouped by gender and then the weighting becomes
irrelevant as we are no longer trying to represent the whole population. 

Summarising by gender will necessarily exclude individuals that are non-binary
or have not stated their gender. In addition, summarising by gender can result 
in (more) categories containing only a few individuals, which is undesirable
as we wish to protect anonymity. Therefore a grouped summary should only be 
used when there is a substantive gender difference and small groups can be 
combined without too much loss of information. First considering the
unweighted and weighted counts can help in making this judgement. Using the 
weights in this exploratory way is perhaps more appropriate than using them to 
adjust the counts, given that they are based in part on predicted data.

## Grouped Summaries

Here we show how the conclusions of the previous section might be applied to 
our illustrative questions.

### Age

The difference in density between weighted and unweighted is small but 
consistent across a wide age range so deserves further exploration. Since we are
using a density summary there is no issue with small sub-groups.

Breaking down by gender shows the difference in age distribution much more 
clearly:

```{r}
data <- na.omit(select(survey_data, Q3, w, Q2)) %>%
  filter(Q2 != "Non-Binary")

ggplot(data, aes(x = Q3, fill = Q2)) + 
  geom_density(alpha = 0.5, color = NA) +
  labs(x = "Year of Birth", y = "Density", fill = NULL, 
       title = "Year of birth of useR! attendees") +
  theme_minimal(base_family = "RobotoCondensed-Regular")
```

### Education level

Although there is some evidence of a more educated male population (weighted
counts shift to higher education levels) the difference is small. In addition
the numbers without university-level education are very small and even combined
with the missing group there is potential that individuals are identified when
broken down by gender. Therefore it is best to keep genders combined here and 
base the final summary on unweighted counts.

```{r}
tidyCat <- function(x, n = NULL, drop = FALSE){
  data <- survey_data %>%
    rename_(x = x)
  # keep n biggest categories
  miss <- "Missing"
  if (!is.null(n)) {
    # if drop, combine with missing
    if (drop){
      miss <- "Other/Missing"
      data <- mutate(data, x = fct_lump(x, n, other = miss))
    } else data <- mutate(data, x = fct_lump(x, n))
  }
  # make missing explicit
  data %>%
    mutate(x = fct_explicit_na(factor(x), na_level = miss))
}
bar <- function(x, xlab = NULL, title = NULL, ...) {
  data <- tidyCat(x, ...)
  
  ggplot(data, aes(x = fct_rev(x))) + 
    geom_bar(aes(y = (..count..)/sum(..count..))) +
    coord_flip() + theme_minimal(base_family = "RobotoCondensed-Regular") +
    labs(x = xlab, y = "", title = title) +
    scale_y_continuous(breaks = seq(0, 1, 0.2), limits = c(0, 1), labels = percent)
}  
bar("Q7", n = 4, drop = TRUE)
```


### Employment status

Here the shift from unweighted counts to weighted counts is larger. Although
there are small groups, these can reasonably be combined into an "Other" 
category.

```{r}
genderBar <- function(x, xlab = NULL, title = NULL, ...) {
  # filter to M/F and compute proportions in categories of x by group
  data <- tidyCat(x, ...) %>%
    filter(Q2 %in% c("Male", "Female")) %>%
    group_by(x, Q2) %>%
    summarise(Count = n()) %>%
    group_by(Q2) %>%
    mutate(x = x,
           Percentage = Count/sum(Count)) %>%
    complete(x, Q2)
  
  ggplot(data, aes(x = fct_rev(x),  y = Percentage, fill = Q2)) + 
    geom_bar(stat = "identity", position = "dodge") +
    coord_flip() + theme_minimal(base_family = "RobotoCondensed-Regular") +
    labs(x = xlab, y = "", title = title, fill = NULL) +
    scale_y_continuous(breaks = seq(0, 1, 0.2), limits = c(0, 1), labels = percent)
}  
genderBar("Q8", "", "Employment Status", n = 5)
```

### Time using R

Again there are material differences between unweighted and weighted counts, 
here the original groups can be left as is:

```{r}
genderBar("Q11", "Time", "Time using R")
```

### Recreational use of R

Here there is a clear difference between unweighted and weighted counts. As only
four people (men) said that they used R primarily as a recreational activity,
this can be combined with the missing category:

```{r}
survey_data <- survey_data %>%
  mutate(Q14 = factor(Q14, labels = c("Both recreation and work",
                                      "Primarily recreation", 
                                      "Primarily work")))
genderBar("Q14", "", "Recreational use of R", n = 2, drop = TRUE)
```

## Conclusions
 
Where there is no material difference between genders, the over-representation
of women in the sample is unimportant. Where there is a substantive difference,
it is interesting to explore this difference in a grouped summary, where again
the over-representation of women becomes irrelevant as males and females are
summarised separately.

Therefore rather than base final summaries on weighted counts, which are based 
in part on predicted data, it is better to use the weighted counts as a guide to
when a grouped summary is appropriate.
